RationalGRL: A Framework for Argumentation and Goal Modeling

### 2018-08-13 Response to reviewers2.txt

===== ORIGINAL COMMENT #3
I think that the paper should compare the proposed argumentation structure at least with the beliefs in GRL. In general, how does this argumentation approach really differ from one based on beliefs and, in particular, belief revision? Assume that an argument is a belief of some stakeholder revised by another argument, which is a belief of some other stakeholder?

===== ORIGINAL REPLY #3

@@

The connection between argumentation and belief revision is interesting, especially from a theoretical point of view, and in knowledge representation and reasoning this is a popular topic of research (see references 5-7 below). However, since studying relations between argumentation and belief revision does not seem directly related to our current framework and seems more interesting from a formal logical point of view, we have left this out of the discussion in our paper, instead choosing to focus on a comparison with beliefs in GRL.

====== REPLY
In our first revision, we already added a brief comparison between arguments in RationalGRL and beliefs in GRL in the related work section (under "Beliefs in GRL"). We have moved these paragraphs to "Goal Modeling" in section 8.1, and further extended and sharpened our discussion and comparison with GRL beliefs. Our main argument, which is now fully in the article, is that in GRL, beliefs are more of an afterthought: they have no formal semantics and it is not possible to further reason with them (i.e. given indicate conflicts between beliefs compute ways in which beliefs can be revised). Furthermore, we briefly discuss belief revision and the possibilities this may offer for future research. We still believe that discussing the relations between formal argumentation from AI and belief revision postulates from AI is not directly suited for the current (RE) audience. A more detailed comparison between RationalGRL and a framework based on belief revision would require us to develop a new belief revision framework connected to GRL and then model, for example, the case study or an extended example using this new belief revision framework for GRL. This would provide enough matrial for a new paper.

### 2018-08-15 validation.tex
1st update section 7

In addition to the empirical case study (Section~\ref{sect:gmas}), which forms the basis of the RationalGRL framework, we also performed a small user evaluation with 16 users. This evaluation concerned mainly with the use of the RationalGRL tool, but more general questions regarding the usefulness of arguments in goal modeling were also asked. We first present the details and results of the evaluation, and then provide a more high-level analysis in Section~\ref{sec:eval-an}.

\subsection{Experimental setup}
The study consists of three parts:
\begin{enumerate}
\item Explanation: We explain to the participants the basics of the RationalGRL development process and the RationalGRL tool.
\item Modeling: We ask the participants to model a summarized discussion in the RationalGRL tool.
\item Survey: We ask some questions about the usability of the RationalGRL tool, and how it relates to other goal modeling languages.
\end{enumerate}

The instructions for these three parts are contained in a single document that we sent out to the participants.\footnote{See \url{www.rationalgrl.com}, page ``Empirical Study''} At the end of the document we provide a link to the survey.\footnote{https://goo.gl/forms/fDSMUnAV20wy7kbY2} The instructions we provide to the users for task 2 can be found in Appendix~\ref{sect:survey}.

Two thirds of the participants had either a PhD or Master degree, 20\% a Bachelor degree, and one respondent responded with ``Other''. The participants turned out to have a relatively high level of experience with IT, with an average rating of 6 years, ranging from 0 to 25 years. They also judged themselves to be quite experienced with early-phase requirements engineering (average of rating 3.2 out of 5), but experience with goal modeling languages was less (rating 1.9 out of 5, although two respondents use goal modeling languages almost weekly). The participants of the study had experience with the following goal modeling languages: No experience (7 users), GRL (2 users), i* (5 users), Tropos (1 user), and KAOS (1 user).

@@

\subsection{Usability of the RationalGRL tool}

There are four questions in the survey which ask users to rate various aspects of the RationalGRL tool on a scale of 1 to 5:
\begin{itemize}
\item[Q1] Was it easy or difficult to get started with the RationalGRL tool?
\item[Q2] Was the details pane (containing details of an element, critical questions, etc) easy or difficult to use?
\item[Q3] Was it easy or difficult to understand the way in which the status of arguments and other elements is determined?
\item[Q4] Did you find it easy or difficult to model the example discussion using RationalGRL?
\end{itemize}

\begin{table*}[t]
\centering
\begin{tabularx}{0.95\textwidth}{l|l|l|l|l|l|l|l}
& very difficult & somewhat difficult & neither difficult nor easy & somewhat easy & very easy & avg. score (1-5) & std. dev.\\
\hline
Q1 & 2 (12.5\%) & 1 (6.3\%) & 6 (37.5\%) & 6 (37.5\% & 1 (6.3\%) & 3.2 & 1.1\\
Q2 & 2 (12.5\%) & 2 (12.5\%) & 4 (25\%) & 8 (50\%) & 0 (0\%) & 3.1 & 1.1\\
Q3 & 0 (0\%) & 3 (20\%) & 4 (26.7\%) & 5 (33.3\%) & 3 (20\%) & 3.3 & 1.4\\
Q4 & 2 (12.5\%) & 5 (31.3\%) & 6 (37.5\%) & 3 (18.8\%) & 0 (0\%) & 2.6 & 1.0
\end{tabularx}
\caption{User scores for the usability of the RationalGRL tool}
\label{table:survey:table1}
\end{table*}

\begin{table*}[t]
\centering
\begin{tabularx}{0.95\textwidth}{l|l|l|l|l|l|l|l}
& very useful & somewhat useful & neither useful nor useless & somewhat useless & very useless & avg. score (1-5) & std. dev.\\
\hline
Q5 & 0 (0\%) & 0 (0.0\%) & 3 (21.4\%) & 8 (57.1\% & 3 (21.4\%) & 4.0 & 0.68\\
Q6 & 0 (0.0\%) & 3 (21.4\%) & 3 (21.4\%) & 6 (42.9\%) & 2 (14.3\%) & 3.5 & 1.0\\
Q7 & 0 (0\%) & 0 (0\%) & 2 (14.3\%) & 7 (50.0\%) & 5 (35.7\%) & 4.2 & 0.7
\end{tabularx}
\caption{User scores for the comparison between RationalGRL and other goal modeling languages}
\label{table:survey:table2}
\end{table*}

@@

\subsection{Comparison with other goal modeling languages}

We asked the users three questions about the extension of RationalGRL on a scale of 1 to 5:
\begin{itemize}
\item[Q5] Do you think the arguments and counterarguments of RationalGRL are a useful extension to standard goal modeling?
\item[Q6] Do you think the critical questions and answers in the details pane are a useful extension to standard goal modeling?
\item[Q7] Do you think the automatic determination of the status of arguments and elements is a useful extension to standard goal modeling?
\end{itemize}

@@

>>>>

In addition to the empirical case study (Section~\ref{sect:gmas}), which forms the basis of the RationalGRL framework, we also performed a user evaluation with 16 users. The objective of this evaluation was to determine whether the users found the additions of RationalGRL (arguments, argumentation semantics, critical questions) useful compared to standard GRL, and whether they found it easier to keep track of and express opinions and beliefs using RationalGRL compared to standard GRL. Regarding the comparison of RationalGRL to GRL, the following two broad hypotheses follow from our main points in this paper:

\begin{itemize}
\item[H1] The additions of RationalGRL -- arguments, critical questions and determination of the status of arguments -- are a useful addition to standard goal modelling in GRL.
\item[H2] The additions of RationalGRL make it easier to express, determine the effect of and communicate to other stakeholders one's opinions and beliefs about a goal model.
\end{itemize}

In addition, we wanted to know whether the RationalGRL tool was difficult or easy to use. Note that we did not aim to conduct a usability study for the RationalGRL tool. Rather, we were interested to know if the tool was clearly very easy or difficult to use, as this could influence the participants ideas about the RationalGRL framework and language -- a very nice and easy to use tool tends to make people more positive twoards a particular modelling language, whereas irritations about a bad or buggy tool will lead to a more negative disposition. This leads to the following general hypothesis:

\begin{itemize}
\item[H3] The usability of the RationalGRL tool is neither (very) easy or (very) difficult. 
\end{itemize}

In the rest of this section, we will describe our experiment in more detail and further specify these hypotheses.

\subsection{Experiment design}

The idea was to have our participants perform a small modelling task, and then ask them what they thought of RationalGRL and the tool. Thus, our experiment consisted of three parts:

\begin{enumerate}
\item Explanation: We explain to the participants the basics of the RationalGRL development process and the RationalGRL tool.
\item Modeling: We ask the participants to model a summarized discussion in the RationalGRL tool.
\item Survey: We present the participants with a survey containing questions about the comparison between RationalGRL and GRL, and questions about the usability of the RationalGRL tool.
\end{enumerate}

The instructions for these three parts are contained in a single document that we sent out to the participants.\footnote{See \url{www.rationalgrl.com}, page ``Empirical Study''} The explanation (part 1) starts with a general explanation of GRL and standard goal modelling similar to the explanation provided in section \ref{{sect:background:grl}, then briefly discusses the additions of RationalGRL (argument, attack, critical questions) and finally the tool is briefly described (similar to Section \ref{sect:tool}).

For part 2, we provided a small part of a transcript and some context information (see Appendix~\ref{sect:survey}) and asked the participants to model this in the tool and take a screenshot of their final model. The idea is that the participants are ``present'' at a discussion about the early-stage requirements and are asked to model the goal model corresponding to these requirements.

For part 3, we provided the participants with a survey\footnote{https://goo.gl/forms/fDSMUnAV20wy7kbY2} . The survey starts with general questions about the participant (experience etc.). It then asks the following questions about the RationalGRL tool, asking the participants to rate various aspects of the tool on a Likert scale of 1 (very difficult) to 5 (very easy):
\begin{itemize}
\item[Q1] Was it easy or difficult to get started with the RationalGRL tool?
\item[Q2] Was the details pane (containing details of an element, critical questions, etc) easy or difficult to use?
\item[Q3] Was it easy or difficult to understand the way in which the status of arguments and other elements is determined?
\item[Q4] Did you find it easy or difficult to model the example discussion using RationalGRL?
\end{itemize}
We also ask open questions about the strengths and weaknesses of the tool, possible improvements and whether the participant thinks the tool could in the future be used in practice.

The second part of the survey concerns the comparison between GRL and RationalGRL and starts with Figure~\ref{fig:example-small}, which models the discussion for part 2 (Appendix~\ref{sect:survey}) in standard GRL, so that the participants can compare their own RationalGRL models and experiences with the standard GRL model. The following questions ask the participants to rate RationalGRL on a Likert scale from 1 (very useless) to 5 (very useful):
\begin{itemize}
\item[Q5] Do you think the arguments and counterarguments of RationalGRL are a useful extension to standard goal modeling?
\item[Q6] Do you think the critical questions and answers in the details pane are a useful extension to standard goal modeling?
\item[Q7] Do you think the automatic determination of the status of arguments and elements is a useful extension to standard goal modeling?
\end{itemize}
The following questions ask the participants to rate RationalGRL vs. standard goal modelling on a Likert scale from 1 (much more difficult) to 5 (much easier):
\item[Q8] Do you think using RationalGRL instead of a standard goal modeling language makes it easier or more difficult to for someone to express beliefs and opinions in a goal model?
\item[Q9] Do you think using RationalGRL instead of a standard goal modeling language makes it easier or more difficult to for someone to determine the effect of beliefs and opinions on the resulting goal model?
\item [Q10] Do you think using RationalGRL instead of a standard goal modeling language makes it easier or more difficult to for someone who is not the original author to understand the goal model? 
\end{itemize}
We conclude with two open questions regarding the strengths and weaknesses of RationalGRL compared to GRL.

\paragraph{Participants}
We asked 16 participants in our network to participate in our experiment. Most of the participants were therefore either employed at the first author's company or staff and (ex-)students from the second author's university. Hence, two thirds of the participants had either a PhD or Master degree, 20\% a Bachelor degree, and one respondent responded with ``Other''. All but one participant had a year or more experience with software development. The average experience was 6.2 years (standard deviation 6.7), with 10 participants having less than 5 years of experience and 6 participants having 8 or more years of experience, with one participant having 25 years of experience. The participants also judged themselves to be quite competent in  early-phase requirements engineering: on average they gave themselves a rating of 3.2 out of 5 (standard deviation 1.3) with half of the participants claiming they were at least ``competent'' (rating 4) or ``very competent'' (rating 5). However, experience with goal modeling languages was markedly less: the average rating was 1.9 (standard deviation 1.3), with 9 participants claiming never to have used a goal modeling language (rating 1), and only two participants displaying regular use (monthly, rating 4, or weekly, rating 5). The participants that had experience with goal modelling languages had mostly used i* (5 users), with 2 users being familiar with GRL and 2 users having used another goal modelling language.

\subsection{Results}

@@

\begin{table*}[t]
\centering
\begin{tabularx}{0.95\textwidth}{l|l|l|l|l|l|l|l}
& very useless & useless & neither useful nor useless & useful & very useful & avg. score (1-5) & std. dev.\\
\hline
Q5 & 0 (0\%) & 0 (0.0\%) & 3 (21.4\%) & 8 (57.1\% & 3 (21.4\%) & 4.0 & 0.68\\
Q6 & 0 (0.0\%) & 3 (21.4\%) & 3 (21.4\%) & 6 (42.9\%) & 2 (14.3\%) & 3.5 & 1.0\\
Q7 & 0 (0\%) & 0 (0\%) & 2 (14.3\%) & 7 (50.0\%) & 5 (35.7\%) & 4.2 & 0.7
\end{tabularx}
\caption{Participant ratings of the usefulness of the additions of RationalGRL}
\label{table:survey:table2}
\end{table*}

\begin{table*}[t]
\centering
\begin{tabularx}{0.95\textwidth}{l|l|l|l|l|l|l|l}
& much more difficult & more difficult & neither more difficult nor easier & easier & much easier & avg. score (1-5) & std. dev.\\
\hline
Q8 & 0 (0\%) & 0 (0.0\%) & 3 (21.4\%) & 8 (57.1\% & 3 (21.4\%) & 4.0 & 0.68\\
Q9 & 0 (0.0\%) & 3 (21.4\%) & 3 (21.4\%) & 6 (42.9\%) & 2 (14.3\%) & 3.5 & 1.0\\
Q10 & 0 (0\%) & 0 (0\%) & 2 (14.3\%) & 7 (50.0\%) & 5 (35.7\%) & 4.2 & 0.7
\end{tabularx}
\caption{Participant ratings of whether the additions of RationalGRL make reasoning about a goal model easier}
\label{table:survey:table2}
\end{table*}

\begin{table*}[t]
\centering
\begin{tabularx}{0.95\textwidth}{l|l|l|l|l|l|l|l}
& very difficult & difficult & neither difficult nor easy & easy & very easy & avg. score (1-5) & std. dev.\\
\hline
Q1 & 2 (12.5\%) & 1 (6.3\%) & 6 (37.5\%) & 6 (37.5\% & 1 (6.3\%) & 3.2 & 1.1\\
Q2 & 2 (12.5\%) & 2 (12.5\%) & 4 (25\%) & 8 (50\%) & 0 (0\%) & 3.1 & 1.1\\
Q3 & 0 (0\%) & 3 (20\%) & 4 (26.7\%) & 5 (33.3\%) & 3 (20\%) & 3.3 & 1.4\\
Q4 & 2 (12.5\%) & 5 (31.3\%) & 6 (37.5\%) & 3 (18.8\%) & 0 (0\%) & 2.6 & 1.0
\end{tabularx}
\caption{User scores for the usability of the RationalGRL tool}
\label{table:survey:table1}
\end{table*}

@@

@@

\subsection{Analysis and discussion}
In order to test our main hypotheses H1, H2 and H3, we have to formulate null hypotheses and alternative hypotheses that can be tested. For H1 and H2, we hypothesize that the participants will, on average, rate the additions of RationalGRL as useful (rating 4) or very useful (rating 5) for Q5, Q6 and Q7, and as making it easier (rating 4) or significantly easier (rating 5) for questions Q8, Q9 and Q10. In other words, we say that our null hypothesis and alternative hypothesis for H1 and H2 are as follows:
\begin{itemize}
[] H1$_{0}$ and H2$_{0}$: average rating for questions Q5-Q10 is 3 or lower.
[] H1$_{a}$ and H1$_{a}$: average rating for questions Q5-Q10 is higher than 3.
\end{itemize}

For hypothesis H3, we do not claim that the usability of the tool is especially easy or difficult: we expect the average rating for questions Q1-Q4 to not be significantly higher or lower than 3 (neither easy or difficult). Thus, our null hypothesis and alternative hypothesis for H3 are as follows:  
\begin{itemize}
[] H3$_{0}$: average rating for questions Q1-Q4 is 3.
[] H3$_{a}$: average rating for questions Q1-Q4 is not 3.
\end{itemize}

Below, we will discuss each of the hypotheses in turn.

\paragraph{Hypothesis H1}
The first analysis concerns hypothesis H1, whether the additions of RationalGRL -- arguments, critical questions and determination of the status of arguments -- are a useful addition to standard goal modelling in GRL. Our hypothesis was that on average the participants would rate the additions as either useful (rating 4) or very useful (rating 5) in question Q4, Q5 and Q6. For each separate question Q4, Q5, Q6, we can thus formulate the following null hypothesis H1$_{0}$ and alternative hypothesis H1$_{a}$.

\paragraph{Hypothesis H2}

\paragraph{Hypothesis H3}
